% ************************** Thesis Acknowledgements *****************************

\begin{acknowledgements}      

Lord receive my deepest gratitude for successful completion of a PhD thesis, for directing my paths and for all your help during this journey. On this intense and at times very enjoyable journey, many people have helped me. The time has come for me to thank them for all the support I received.
				
First and foremost, my deepest gratitude goes to Wessel van Wieringen and Saskia Wilting for their insightful guidance, sincere criticism, patient support and all the efforts and time that they put in me. They offered everything that a student could ask for from his or her advisor. I cannot imagine how I could accomplish my dissertation without their guidance and help. They also left an enduring imprint on me with their way of approaching work and research, which I will benefit from for my entire career. Words alone cannot express the respect and gratitude my family and I do have for you.
					
I would also like to express my gratitude towards my promoters Mark van de Wiel and Renske Steenbergen, as well as to Peter Snijders, for tremendous support, giving valuable suggestions that improved my dissertation.  It is incredibly sad that Peter Snijders is no longer with us to witness the completion of this project.
				
I wish to thank the members of the reading committee, including Dr Hans Berkhof, Prof Dr Mathisca de Gunst, Prof Dr Jeanine Houwing-Duistermaat, Prof Dr Ed Schuuring, Prof Dr Ir Wim van Criekinge for spending their valuable time on careful reading of my thesis and giving me interesting feedback.
				
I would like to thank past and present members of the Epidemiology $\&$ Biostatistics Department  and Pathology Department at the Amsterdam University Medical Centers, that I have had the pleasure to work with. The group has been a source of friendship, good advice and collaboration. I thank particularly to Nimisha Chaturvedi and Gwenael Leday for all their explanations and help during tough times I had in the beginning of my PhD. I thank to Annelieke Jaspers, who performed all the experiments, as well as to Iris Babion for the validation of the experiments and helping with the biological part. I am grateful to the: Gino Kpogbezan, Carel Peeters and Renee de Menzes for their advice and support. Also, I am grateful to Kristian Miok and Andrea Bassi my paranimphs.
				
I am extremely grateful to father Meletios Webber. Thank you for everything you have done for me, for all your prayers, advices and support during the tough times.
				
%Drag\u{a} familia mea vreau s\u{a} va mul$\cb{t}$umesc pentru toat\u{a} jertfa si rug\u{a}ciunile voastr\u{a} ca \cb{s}i a \^{i}nainta\cb{s}ilor no\cb{s}tri. Tat\u{a}, i\cb{t}i mul\cb{t}umesc c\u{a} mai sus\cb{t}inut, c\u{a} ai crezut in mine de la primul meu g\u{a}nd ca s\u{a} plec pe drumul acesta, c\u{a} te-ai dedicat cu totul si c\u{a} mai ajutat la fie care pas. Mam\u{a}, i\cb{t}i mul\cb{t}umesc pentru toat\u{a} grija care ai avut pentru noi to\cb{t}i \cb{s}i pentru intelegerea \cb{s}i ajutorul t\u{a}u. Cristi, i\cb{t}i mul\cb{t}umesc, c\u{a} ai \cb{s}tiut cum sa m\u{a} sustini meru, duc\u{a}nd greuta\cb{t}ile mele \cb{s}i accept\u{a}ndu m\u{a} a\cb{s}a cum sunt. Livija, i\cb{t}i mul\cb{t}umesc pentru coperta c\u{a}r\cb{t}i \cb{s}i mai ales pentru r\u{a}bdarea, ajutorul \cb{s}i dragostea ta. Dumnezeu s\u{a} va ajute!

Drag\u{a} familia mea, vreau s\u{a} v\u{a} mul\c{t}umesc pentru toat\u{a} jertfa \c{s}i rug\u{a}ciunile voastre ca \c{s}i a \^{i}nainta\c{s}ilor no\c{s}tri. Tat\u{a}, \^{i}\c{t}i mul\c{t}umesc c\u{a} m-ai sus\c{t}inut, c\u{a} ai crezut \^{i}n mine de la primul meu g\^{a}nd de a pleca pe drumul acesta, c\u{a} te-ai dedicat cu totul \c{s}i c\u{a} m-ai ajutat la fiecare pas. Mam\u{a}, \^{i}\c{t}i mul\c{t}umesc pentru toat\u{a} grija pe care ai avut-o pentru noi to\c{t}i \c{s}i pentru \^{i}ntelegerea \c{s}i ajutorul t\u{a}u. Cristi, \^{i}\c{t}i mul\c{t}umesc c\u{a} ai \c{s}tiut cum s\u{a} m\u{a} sus\c{t}ii mereu, duc\^{a}nd greut\u{a}\c{t}ile mele \c{s}i accept\^{a}ndu-m\u{a} a\c{s}a cum sunt. Livia, \^{i}\c{t}i mul\c{t}umesc pentru coperta c\u{a}r\c{t}ii \c{s}i mai ales pentru r\u{a}bdarea, ajutorul \c{s}i dragostea ta. Dumnezeu s\u{a} v\u{a} ajute!
\par \begin{flushright} Viktorian Miok\\
Belgrade, April 2018 \end{flushright}
\end{acknowledgements}
