
\chaptermark{Biography}

\begin{biography}


{\bf Viktorian Miok} was born on October 27, 1985 in Zrenjanin, Serbia. In 2004 he graduated from high school Zrenjaninska Gimnazia in Zrenjanin. Viktorian studied Mathematics at the University of Belgrade, Serbia. He graduated in 2008, while in the same year he taught mathematics at the Economical High school in Zrenjanin. He received the Master of Science degree in Applied Statistics and Informatics from West University of Timisoara, Romania in 2010. During this study, he has also been employed as SAS Programmer at the Contact Research Organization Cmed in Timisoara. In the same year he has been given an opportunity to work as research fellow in statistics at the University of Bucharest, Romania under the supervision of Prof M Dumitrescu. In January 2012 he started his doctoral studies in statistics for integrative bioinformatics on joint project between the Department of Epidemiology $\&$ Biostatistics (Prof Dr MA van de Wiel) and Department of Pathology (Prof Dr PJF Snijders) of the Amsterdam University Medical Centers, which resulted in this thesis. Currently he is employed as Bioinformatician at Seven Bridges Genomics in Belgrade. 

\end{biography}



%\chapter{Biography}
%\label{chapter:biography}

%\graphicspath{{Biography/Figs/}{Biography/Figs/PDF/}{Biography/Figs/}}%

%{\bf Viktorian Miok} was born on October 27, 1985 in Zrenjanin, Serbia. In 2004 he graduated from high school Zrenjaninska Gimnazia in Zrenjanin. Viktorian studied Mathematics at the University of Belgrade, Serbia. He graduated in 2008, while in the same year he taught mathematics at the Economical High school in Zrenjanin. He received the Master of Science degree in Applied Statistics and Informatics from West University of Timisoara, Romania in 2010. During this study, he has also been employed as SAS Programmer at the Contact Research Organization Cmed in Timisoara. In the same year he has been given an opportunity to work as research fellow in statistics at the University of Bucharest, Romania under the supervision of prof. dr. M Dumitrescu. In January 2012 he started his doctoral studies in statistics for integrative bioinformatics on joint project between the Department of Epidemiology $\&$ Biostatistics (prof. dr. ir. MA van de Wiel) and Department of Pathology (prof. dr. PJF Snijders) of the VU University Medical Center in Amsterdam, which result in this thesis. Currently he is employed as Bioinformatics Analyst at Seven Bridges Genomics in Belgrade. 


% ******************************* Thesis Conclusion ********************************